---
title: "Hands-on Exercise 1 - A Layered Grammar of Graphics: ggplot2 methods"
---

## Learning Outcome

-   basic principles and essential components of ggplot2

-   plot statistical graphics based on the principle of Layered Grammar of Graphics

-   apply the essential graphical elements provided by ggplot2 to create elegant and yet functional statistical graphics.

## Getting started

### Install and load required libraries

Before we get started, it is important for us to ensure that the required R packages have been installed. If yes, we will load the R packages. If they have yet to be installed, we will install the R packages and load them onto R environment.

```{r}
pacman::p_load(tidyverse)
```

### **Importing data**

-   The code chunk below imports *exam_data.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package.

-   **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("data/Exam_data.csv", show_col_types = FALSE)
```

-   Year end examination grades of a cohort of primary 3 students from a local school.

-   There are a total of seven attributes. Four of them are categorical data type and the other three are in continuous data type.

    -   The categorical attributes are: ID, CLASS, GENDER and RACE.

    -   The continuous attributes are: MATHS, ENGLISH and SCIENCE.

## **Introducing ggplot**

ggplot2 is an R package for **declaratively** creating **data-driven** graphics based on ***The Grammar of Graphics.***

It is also part of the tidyverse family specially designed for visual exploration and communication. For more detail, visit [ggplot2 link](https://ggplot2.tidyverse.org/).

### **R Graphics VS ggplot**

First, let us compare how R Graphics, the core graphical functions of Base R and ggplot plot a simple histogram.

R Graphics:

```{r}
hist(exam_data$MATHS)
```

ggplot2:

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

If the code chunk is relatively simple if R Graphics is used, then why is ggplot2 recommended?

## **Grammar of Graphics**

Grammar of Graphics is a general scheme for data visualization which breaks up graphs into semantic components such as scales and layers. It defines the rules of structuring mathematical and aesthetic elements into a meaningful graph.

There are two principles in Grammar of Graphics, they are:

-   Graphics = distinct layers of grammatical elements

-   Meaningful plots through aesthetic mapping

### **A Layered Grammar of Graphics**

ggplot2 is an implementation of the Grammar of Graphics.

The seven grammars of ggplot2 are:

1.  **Data**: The dataset being plotted.

2.  **Aesthetics** take attributes of the data and use them to influence visual characteristics, such as position, colours, size, shape, or transparency.

3.  **Geometrics**: The visual elements used for our data, such as point, bar or line.

4.  **Facets** split the data into subsets to create multiple variations of the same graph (paneling, multiple plots).

5.  **Statistics**, statiscal transformations that summarise data (e.g. mean, confidence intervals).

6.  **Coordinate systems** define the plane on which data are mapped on the graphic.

7.  **Themes** modify all non-data components of a plot, such as main title, sub-title, y-aixs title, or legend background.

## **Essential Grammatical Elements in ggplot2: data**

```{r}
ggplot(data=exam_data)
```

Observations:

-   A blank canvas appears.

-   `ggplot()` initializes a ggplot object.

-   The *data* argument defines the dataset to be used for plotting.

-   If the dataset is not already a data.frame, it will be converted to one by `fortify()`.

## **Essential Grammatical Elements in ggplot2: [Aesthetic mappings](https://ggplot2.tidyverse.org/articles/ggplot2-specs.html)**

The aesthetic mappings take attributes of the data and and use them to influence visual characteristics, such as position, colour, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics of a plot are specified in the [`aes()`](https://ggplot2.tidyverse.org/reference/aes.html) function call (in later part of this lesson, you will see that each *geom* layer can have its own aes specification)

Code chunk below adds the aesthetic element into the plot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

Observation: ggplot includes the x-axis and the axis label.

## **Essential Grammatical Elements in ggplot2: geom**

Geometric objects are the actual marks we put on a plot. Examples include:

-   *geom_point* for drawing individual points (e.g., a scatter plot)

-   *geom_line* for drawing lines (e.g., for a line charts)

-   *geom_smooth* for drawing smoothed lines (e.g., for simple trends or approximations)

-   *geom_bar* for drawing bars (e.g., for bar charts)

-   *geom_histogram* for drawing binned values (e.g. a histogram)

-   *geom_polygon* for drawing arbitrary shapes

-   *geom_map* for drawing polygons in the shape of a map! (You can access the data to use for these maps by using the map_data() function).

-   *geom_col*

-   *geom_boxplot*

-   *geom_density*

```{=html}
<!-- -->
```
-   A plot must have at least one geom; there is no upper limit. You can add a geom to a plot using the **+** operator.

-   For complete list, please refer to [here](https://ggplot2.tidyverse.org/reference/#section-layer-geoms).

### **Geometric Objects: geom_bar**

The code chunk below plots a bar chart by using [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

### **Geometric Objects: geom_dotplot**

In a dot plot, the width of a dot corresponds to the bin width (or maximum width, depending on the binning algorithm), and dots are stacked, with each dot representing one observation.

In the code chunk below, [`geom_dotplot()`](https://ggplot2.tidyverse.org/reference/geom_dotplot.html) of ggplot2 is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

Note that the above y scale is not very useful, in fact it is very misleading.

The code chunk below performs the following two steps:

-   `scale_y_continuous()` is used to turn off the y-axis, and

-   *binwidth argument* is used to change the binwidth to 2.5.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

### **Geometric Objects: `geom_histogram()`**

In the code chunk below, [*geom_histogram()*](https://ggplot2.tidyverse.org/reference/geom_histogram.html) is used to create a simple histogram by using values in *MATHS* field of *exam_data*.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

Note that the default bin is **30**.

### **Modifying a geometric object by changing `geom()`**

In the code chunk below,

-   *bins* argument is used to change the number of bins to 20,

-   *fill* argument is used to shade the histogram with light blue color, and

-   *color* argument is used to change the outline colour of the bars in black

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

### **Modifying a geometric object by changing *aes()***

-   The code chunk below changes the interior colour of the histogram (i.e. *fill*) by using sub-group of *aesthetic()*.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

### **Geometric Objects: `geom-density()`**

[`geom-density()`](https://ggplot2.tidyverse.org/reference/geom_density.html) computes and plots [kernel density estimate](https://en.wikipedia.org/wiki/Kernel_density_estimation), which is a smoothed version of the histogram.

It is a useful alternative to the histogram for continuous data that comes from an underlying smooth distribution.

The code below plots the distribution of Maths scores in a kernel density estimate plot:

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

The code chunk below plots two kernel density lines by using *colour* or *fill* arguments of *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           fill = GENDER)) +
  geom_density()
```

### **Geometric Objects: geom_boxplot**

[`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html) displays continuous value list. It visualises five summary statistics (the median, two hinges and two whiskers), and all “outlying” points individually.

The code chunk below plots boxplots by using [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_boxplot.html).

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

[**Notches**](https://sites.google.com/site/davidsstatistics/home/notched-box-plots) are used in box plots to help visually assess whether the medians of distributions differ. If the notches do not overlap, this is evidence that the medians are different.

The code chunk below plots the distribution of Maths scores by gender in notched plot instead of boxplot:

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

### **Geometric Objects: `geom_violin`**

[`geom_violin`](https://ggplot2.tidyverse.org/reference/geom_violin.html) is designed for creating violin plot. Violin plots are a way of comparing multiple data distributions. With ordinary density curves, it is difficult to compare more than just a few distributions because the lines visually interfere with each other. With a violin plot, it’s easier to compare several distributions since they’re placed side by side.

The code below plot the distribution of Maths score by gender in violin plot:

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

### **Geometric Objects: `geom_point()`**

[`geom_point()`](https://ggplot2.tidyverse.org/reference/geom_point.html) is especially useful for creating scatterplot.

The code chunk below plots a scatterplot showing the Maths and English grades of pupils by using `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

### ***geom*** **objects can be combined**

The code chunk below plots the data points on the boxplots by using both `geom_boxplot()` and `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## **Essential Grammatical Elements in ggplot2: stat**

The [Statistics functions](https://ggplot2.tidyverse.org/reference/#stats) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)

    -   a mean

    -   a confidence limit

-   There are two ways to use these functions:

    -   add a `stat_()` function and override the default geom, or

    -   add a `geom_()` function and override the default stat.

### **Working with `stat()`**

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

### **Working with stat - the *stat_summary()* method**

The code chunk below adds mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun.y="mean",         
               colour ="red",        
               size=4)               
```

### **Working with stat - the `geom()` method**

The code chunk below adding mean values by using `geom_()` function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun.y="mean",           
             colour ="red",          
             size=4)          
```

### **Adding a best fit curve on a scatterplot?**

The scatterplot below shows the relationship of Maths and English grades of pupils. The interpretability of this graph can be improved by adding a best fit curve.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()  
```

In the code chunk below, [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5)
```

## **Essential Grammatical Elements in ggplot2: Facets**

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) and [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

### **Working with `facet_wrap()`**

[`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

The code chunk below plots a trellis plot using `facet-wrap()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### **`facet_grid()` function**

[`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

The code chunk below plots a trellis plot using `facet_grid()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## **Essential Grammatical Elements in ggplot2: Coordinates**

The *Coordinates* functions map the position of objects onto the plane of the plot. There are a number of different possible coordinate systems to use, they are:
