---
title: "In-class Exercise 1"
author: "Kwok Pei Shan"
date: "April 13, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

[Notes:]{.underline}

-   eval: true --\> code chunks will be run

-   echo: true --\> code chunks will show in webpage

-   warning: false --\> will not show warning message

-   freeze: true --\> will not render if there is no change to qmd file

what's the difference between code block and code chunk?

## Loading R packages

In the code chunk below, [`p_load()`](https://www.rdocumentation.org/packages/pacman/versions/0.5.1/topics/p_load) of [pacman](http://trinker.github.io/pacman_dev/) package is used to load tidyverse family of packages.

Note: If it's used only once - instead of loading the whole pacman package, we can do this instead to consume less resources:

```{r}
pacman::p_load(tidyverse)

# Advantage of using p_load:
# you can load multiple packages in one line!
```

Instead of using p_load, you can also do this:

```{r}
#| eval: false
library(tidyverse)
#library(...)

# But by doing this, have to create a new line for each package that I need to load.
```

## Loading Data

Method 1 (preferred method):

```{r}
realis <- read_csv("data/realis2019.csv", show_col_types = FALSE)

# read_csv is from tidyverse package
# class(realis) is a tibblr dataframe
# original column names will be retained.

```

Method 2:

```{r}
realis.csv <- 
  read.csv("data/realis2019.csv")

# read.csv is from Base R package
# class(realis.csv) is a dataframe.
# column names will be affected (e.g. spaces will be replaced with dots)

```

Advantage of using R over Python is that R already has base statistical packages, e.g., can just use hist() to plot histogram using the Base R package.

The code chunk below uses ggplot package to plot histogram:

```{r}
ggplot(data = realis,
       aes(x = `Unit Price ($ psm)`)) +
  geom_histogram()
```
